---
title: "Digital Exercise: Correct or Incorrect"
subtitle: "Final Project for Practical ML Course from Data Science Specialization"
author: "Ivo Georgiev"
date: "August 10, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

TODO: Overall parameters are:
- 1999 words
- 4 figures
- gh-pages branch (to view HTML directly in the repository for grading)


## Abstract

## Exploratory Analysis

```{r explore-1}
pmlTrain <- read.csv("pml-training.csv", na.strings = c("", "NA"))
pmlTest <- read.csv("pml-testing.csv", na.strings = c("", "NA"))
dim(pmlTrain)[[2]]
c(dim(pmlTrain)[[1]], dim(pmlTest)[[1]])
```

There are 19622 training samples and 20 testing samples. There are 160 columns. The predicted variable is __classe__ and it represents one of 5 classes of performing a _unilateral dumbbell biceps curl_. It takes values "A", "B", "C", "D", and "E". The meanings are defined on the [HAR website](http://groupware.les.inf.puc-rio.br/har) in the [QAR paper](http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf). In short, "A" represents an exercise performed according to specification, while the other classes represent particular identifiable errors in the performance of the exercise.

The following shows the number of training samples for each class (__classe__):

```{r explore-2}
aggregate(pmlTrain$classe, list(pmlTrain$classe), length)
```

Looking at the columns of the train set...

```{r explore-3, echo=FALSE}
trainAllNA <- sum(sapply(pmlTrain, function(x) all(is.na(x))))
trainNoNA <- sum(sapply(pmlTrain, function(x) all(!is.na(x))))
trainSomeNA <- sum(sapply(pmlTrain, function(x) !all(!is.na(x))))
cols <- c(trainAllNA, trainNoNA, trainSomeNA)
names(cols) <- c("All NA", "No NA", "Some NA")
cols
```

and the test set...

```{r explore-4, echo=FALSE}
trainAllNA <- sum(sapply(pmlTest, function(x) all(is.na(x))))
trainNoNA <- sum(sapply(pmlTest, function(x) all(!is.na(x))))
trainSomeNA <- sum(sapply(pmlTest, function(x) !all(!is.na(x))))
cols <- c(trainAllNA, trainNoNA, trainSomeNA)
names(cols) <- c("All NA", "No NA", "Some NA")
cols
```

We see that the training set contains no empty (all NA) columns, 60 columns w/o missing values, and 100 columns with some missing values. The test set has 100 empty columns and 60 complete columns.

