---
title: "Is Automatic or Manual Transmission Better for MPG"
subtitle: "Final Project for Regression Models Course from Data Science Specialization"
author: "Ivo Georgiev"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TODO

1. Follow the approach in ISLR 3.2.2 and 3.4, following the 4 (7 in 3.4) questions:
    + Is at least one of the predictors X1 , X2 , . . . , Xp useful in predicting the response?
    + Do all the predictors help to explain Y, or is only a subset of the predictors useful?
    + How well does the model fit the data?
    + Given a set of predictor values, what response value should we predict, and how accurate is our prediction?
    
2. Revisit intercept vs no-intercept. If intercept, interpret it correctly, i.e. explain _what is the baseline_.
3. Plots:
    + before and after _adjustment_
    + confidence/prediction intervals
    + multi-dimensional plot highlighting the residuals, potentially testing the linearity assumption (to help with (4) below)
    + ggplot2 outside of exploratory analysis
4. Investigate relaxations of the linear model assumptions, additivity and linearity, and interpret in the conclusion.
    + interaction terms of transmission type with other variables
    + investigate and plot two parallel lines vs interaction term (see ISLR Fig 3.7)
    + polynomial regression for the quantitative adjustment regressors (have to investigate and plot separately)
5. Investigate the potential problems in ISLR 3.3.3 (also see [Quick-R](http://www.statmethods.net/stats/rdiagnostics.html)):
    + Non-linearity of the response-predictor relationships
    + Correlation of error terms
    + Non-constant variance of error terms
    + Outliers
    + High-leverage points
6. Collinearity.
    + collinearity (plot) vs correlation (number)
    + multi-collinearity and the variance inflation factor (VIF)
7. Explore the contrasts
8. Rewrite the abstract to reflect the new analysis. State everything clearly.
9. Style:
    + remove verbosity
    + leave enough r code chunks in the main section to keep the document together
    + what to do with r code chunk output: intersperse into main text or format as appendix figures?
    + consistent and thorough figure captions
    + to save figure space, collect related plots into panels

## Abstract

This document is a short study of the __mtcars__ dataset to determine the effect on car mileage per gallon (MPG) of automatic and manual transmission. With 95% confidence, we estimate that a car with manual transmision will have between 3 and 11 mpg higher mileage than an _equivalent_ car with automatic transmission. However, this small data set provides _insufficient_ evidence to confidently infer that manual transmission correlates with better performance in the wider population of cars.

## Exploratory Analysis

A quick look look at the __mtcars__ dataset (see Appendix A) shows that:

1. This is a small dataset, with few observations relative to regressors, necessitating the use of T tests and statistics.
2. The following variables should be treated as factors: the variable of interest __am__ (type of transmission: 0 - automatic, 1 - manual), and also __cyl__ (number of cylinders), __vs__ (type of engine: 0 - __V__-engine, 1 - **s**traight), __gear__ (number of gears), and __carb__ (number of carburetors). The transformed dataset is called __cars__, shown in Fig A.1.
3. Rows represent cars of different makes and classes, so the samples are independent, but there is little information on how the data was collected and therefore little ground for futher assumptions about the sample distribution and variance.

```{r cars, echo=FALSE, results='hide'}
cars <- transform(mtcars, cyl=as.factor(cyl), vs=as.factor(vs), gear=as.factor(gear), am=as.factor(am), carb=as.factor(carb))
```

A quick plot of _MPG vs type of transmission_ (Fig A.2) suggests that manual transmission is better for MPG, though:

1. The dataset is _small_.
2. There is quite a bit of _variation_ and _overlap_.
3. The data subset for manual transmission appears _skewed_.

Another way to show the same effect is to use a T test of MPG for the samples with manual and automatic transmission:

```{r, ttest, echo=FALSE}
t<-t.test(cars[cars$am=="0",]$mpg, cars[cars$am=="1",]$mpg, paired=FALSE, var.equal=FALSE)
s<-c(t$statistic,t$p.value,t$conf.int)
names(s)<-c("t statistic", "p-value", "Conf low", "Conf high")
s
```

Indeed, according to this test  we can confidently conclude that 95% of the time a car with manual transmission will have higher MPG than a car with automatic transmission. Fitting a linear model can allow us to quantify this effect:

```{r, lm-am-only, echo=FALSE}
summary(lm(mpg~am-1, cars))$coefficients
```

So, we can reasonably expect that 95% of the time a car with manual transmission will have on average 7 mpg higher mileage than a car with automatic transmission. (See Fig A.3).

What this is saying, is that _all else being equal_ a car with manual transmission will drive 7 more miles per gallon than the __same__ car with automatic transmission. However, the dataset _does NOT support_ such a conclusion, since it clearly violates the assumption. We can see this when we add more regressors to our simple model.

## Statistical Analysis

The dataset has only two rows which allow us to compare the transmission type effect on MPG for _similar_ vehicles, the Corola and the Corona:

```{r toyota, echo=FALSE}
cars[grep("^Toyota", rownames(cars)),]
```

Even here, we see that the weight, engine size, and horsepower can __confound__ the inference about effect of transmission type on car mileage. How to pick the additional regressors? Looking at the correlation matrix for the non-factor variables in the dataset (see Appendix B), we see that weight (__wt__), horsepower (__hp__), and engine displacement (__disp__) have the highest correlation with MPG. However, horsepower is highly correlated with displacement (i.e. engine size), so only one of the two should be used. Adding weight and displacement to our model, we see that the effect of manual vs automatic transmission on MPG is almost entirely wiped out:

```{r confound, echo=FALSE}
summary(lm(mpg~am+disp+wt-1,cars))$coefficients
```

Investigating the nested model variance for the rest of the regressors (see Appendix B), we see that only engine type (__vs__) might be useful to include in the model. Adding this regressor gives us our final model:

```{r final, echo=FALSE}
summary(lm(mpg~am+disp+wt+vs-1,cars))$coefficients
```

We see that the effect of transmission type cannot be isolated easily from the rest of the factors that determine mileage performance. Manual transmission only improves mileage by about 1.5 mpg relative to automatic.

## Conclusion

Cars are complicated machines and many factors affect the performance criteria. Cars are made for different purposes, from commuting to racing, and are designed and tuned accordingly. The __mtcars__ dataset is extremely small but at the same time very variable, making it difficult to determine the effect of transmission type on MPG. There are no data that allow the comparison of otherwise similar or equivalent cars with manual and automatic transmission and making a strong inference on the effect of transmission type on mileage. It is a safe bet that the relation is very complex and highly non-linear.

\pagebreak

## Appendix

### A. Exploratory Analysis

```{r explore-1, echo=FALSE, results='hide'}
require(datasets)
data(mtcars)
```

```{r explore-2, echo=FALSE, results='hide'}
str(mtcars)
dim(mtcars)
sapply(mtcars, unique)
```

```{r explore-3, echo=FALSE}
plot(cars, main="Fig A.1: Plot of the cars dataset, showing factors")
```

```{r explore-4, echo=FALSE}
boxplot(mpg~am, data=cars, main="Fig A.2: MPG by transmission type", col=c("red", "blue"), xaxt="n")
axis(1, at = 1:2, labels=c("Automatic", "Manual"))
```

```{r explore-5, echo=FALSE}
par(mfrow=c(2,2), oma=c(2,0,0,0))
plot(lm(mpg~am-1,cars))
mtext("Fig A.3: Linear model of MPG vs transmission type", side=1, outer = TRUE)
```

```{r explore-6, echo=FALSE, results='hide'}
dev.off() # reset the graphical device
```

### B. Statistical Analysis

Correlation table for the non-factor variables in the dataset:

```{r stat-1, echo=FALSE}
cor(cars[c("mpg", "disp", "hp", "drat", "wt", "qsec")])
```

Analysis of variance for nested models for the dataset:

```{r stat-2}
fit1 <- lm(mpg~am+disp+wt-1, cars)        # initial confounding model
fit2 <- update(fit1, mpg~am+disp+wt+vs-1) # final model
fit3 <- update(fit1, mpg~am+disp+wt+vs+cyl-1)
fit4 <- update(fit1, mpg~am+disp+wt+vs+cyl+qsec-1)
fit5 <- update(fit1, mpg~am+disp+wt+vs+cyl+qsec+gear-1)
fit6 <- update(fit1, mpg~am+disp+wt+vs+cyl+qsec+gear+carb-1)
anova(fit1,fit2,fit3,fit4,fit5,fit6)
```

```{r stat-3, echo=FALSE}
par(mfrow=c(2,2), oma=c(2,0,0,0))
plot(lm(mpg~am-1,cars))
mtext("Fig B.1: Confounded model of MPG vs transmission type", side=1, outer = TRUE)
```
