---
title: "Is Automatic or Manual Transmission Better for MPG"
subtitle: "Final Project for Regression Models Course from Data Science Specialization"
author: "Ivo Georgiev"
date: "August 9, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

TODO - write last, in the style of the following sentences from [here](http://simplystatistics.org/2013/01/28/data-supports-claim-that-if-kobe-stops-ball-hogging-the-lakers-will-win-more/)

“Data supports the claim that if Kobe stops ball hogging the Lakers will win more”
“Linear regression suggests that an increase of 1% in % of shots taken by Kobe results in a drop of 1.16 points (+/- 0.22) in score differential.”

## Exploratory Analysis

TODO - label appendix figures and reference them in main text

Taking a quick look at the __mtcars__ dataset shows that:

1. This is a small dataset with very few observations, necessitating T tests and statistics.
2. The following variables should be treated as factors: __cyl__ (number of cylinders), __vs__ (type of engine: 0 - __V__-engine, 1 - **s**traight), __am__ (type of transmission: 0 - automatic, 1 - manual), __gear__ (number of gears), and __carb__ (number of carburetors).

A quick plot of _MPG vs type of transmission_ suggests that manual transmission is better for MPG, though:

1. The dataset is _small_.
2. There is quite a bit of _variation_ and _overlap_.
3. The data subset for manual transmission appears _skewed_.

These points should be investigated to quantify the effect of transmission on MPG. 

## Statistical Analysis

TODO:
- confounders
- missing variables (different makes/classes, driving style and idling/speeding/cruising, tuning for different classes and target clientelle)
- am + all
- am + all - gear
- am + gear
- am + drat
- am + hp
- am + wt
- am + carb
- am within a make, within a class
- for each, check the dataset to see what values for am and the other variable actually exist (this will diminish the data significantly)

\pagebreak

## Appendix

TODO - how to format R cmd line output as a figure (e.g. __str(mtcars)__)? caption?

### A. Exploratory Analysis

```{r explore-1, echo=FALSE, results='hide'}
require(datasets)
data(mtcars)
```

```{r explore-2, echo=FALSE}
str(mtcars)
```

```{r explore-3, echo=FALSE}
sapply(mtcars, unique)
```

### B. Statistical Analysis


## References

TODO
ref for dataset [SO link](http://stackoverflow.com/questions/18617174/r-mtcars-dataset-meaning-of-vs-variable)
